---
title: "htseq-deseq2"
author:
    - "RNA-auto"
date: "`r format(Sys.time(), '%d %B, %Y')`"
params:
   rmd: "salmon-deseq.Rmd"
output:
  html_document:
  highlight: tango
  number_sections: no
  theme: default
  toc: yes
  toc_depth: 3
  toc_float:
    collapsed: no
    smooth_scroll: yes
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# DESeq2 analysis on HTSeq output

## Load libraries
```{r load_libraries, message=FALSE, warning=FALSE}
library("DESeq2")
library("vsn")
library("pheatmap")
library("EnhancedVolcano")
library("apeglm")
library("ggplot2")
library("ashr")
library("gridExtra")
library("grid")
library("GenomicFeatures")
library("rtracklayer")
```

## Get I/O from snakemake

```{r get_io, message=FALSE, warning=FALSE}
# Load the count data and the meta data for the samples

out.html <- as.character(snakemake@output)
out.list <- paste0(dirname(out.html), "/gene_list.tsv")

gene.db <- as.character(snakemake@input[["gene_db"]])

count_dir <- as.character(snakemake@params[["count_dir"]])

col_data_file <- as.character(snakemake@input[["sample_data"]])

colData <- read.table(col_data_file, header = TRUE)
colData$group <- sub("-","_",colData$group)

sampleFiles <- grep("counts.tsv", list.files(count_dir), value=TRUE)
sampleNames <- sub("_counts.tsv", "", sampleFiles)

sampleTable <- data.frame(sampleName = sampleNames,
                          fileName = sampleFiles,
                          condition = colData$condition,
                          group = colData$group)
```

## Load genes from gtf
```{r load_genes, message=FALSE, warning=FALSE}
library("ensembldb")
EDB <- EnsDb(gene.db)
ensembl.genes <- genes(EDB)
seqlevels(ensembl.genes) <- paste0("chr", seqlevels(ensembl.genes))
```

```{r assemble_data, message=FALSE, warning=FALSE}
# Make the count data into a dataframe that can be passed on to DESeq2
dds <- DESeqDataSetFromHTSeqCount(sampleTable = sampleTable,
                                  directory = count_dir,
                                  design= ~ condition)


#keep <- apply(counts(dds), 1, function(x){all(x > 5) && sum(x) > 150 } )
keep <- apply(counts(dds), 1, function(x){sum(x) > 150 } )
dds <- dds[keep,]
dds$group <- factor(dds$group)
dds$condition <- factor(dds$condition, levels=c("A","C"))

```

## Run DESeq2
```{r run_deseq2, message=FALSE, warning=FALSE}
# Run DESeq
dds <- DESeq(dds)
```

## Plot standard deviation of genes over mean

### For log2 normalized counts
```{r ntd_plot, message=FALSE, warning=FALSE}
ntd <- normTransform(dds)
vsd <- vst(dds)
meanSdPlot(assay(ntd))
```
### For vst transformed data

```{r vsd_plot, message=FALSE, warning=FALSE}
meanSdPlot(assay(vsd))
```

## Plot the estimated dispersion of the genes over mean
```{r plot_disp, message=FALSE, warning=FALSE}
plotDispEsts(dds)
```

## heatmap for the samples vs the 500 genes with highest variance

```{r heatmap, message=FALSE, warning=FALSE}
select <- order(rowVars(counts(dds,normalized=TRUE), decreasing=TRUE),
               decreasing=TRUE)[1:500]
df <- as.data.frame(colData(dds)[,c("group", "condition")])
DAT <- assay(vsd)[select,]
rownames(df) <- colnames(DAT)
pheatmap(DAT, cluster_rows=TRUE, show_rownames=FALSE,
         cluster_cols=TRUE, annotation_col=df)
```

## PCA
```{r pca, message=FALSE, warning=FALSE}
library("PCAtools")

metadata <- data.frame(condition=dds$condition)
metadata$group <- dds$group
row.names(metadata) <- colnames(assay(vsd))


p <- pca(assay(vsd), metadata = metadata, removeVar = 0)
biplot(p, colby="group", shape= "condition", legendPosition = "top")
pairsplot(p, colby="condition", components=getComponents(p, c(1:5)), margingaps = unit(c(-0.01, -0.01, -0.01, -0.01), "cm"))
```

## Results

```{r get_results, message=FALSE, warning=FALSE}
conditions <- unique(as.character(dds$condition))
conditions <- paste0("condition_", conditions[2],"_vs_", conditions[1])
res <- results(dds, name=conditions)
res <- lfcShrink(dds, coef=conditions, res=res, type = 'apeglm')
```

### Volcano plots
```{r volcano, message=FALSE, warning=FALSE}
pCutoff = 10e-10
EnhancedVolcano(res,
    lab = rownames(res),
    x = 'log2FoldChange',
    y = 'pvalue',
    xlim = c(-10, 10),
    title = 'A versus C; apeglm shrunken lfc',
    pCutoff = pCutoff,
    FCcutoff = 2,
    pointSize = 3.0,
    labSize = 3.0)
```
zoomed in volcano

```{r volcano_zoom, message=FALSE, warning=FALSE}
EnhancedVolcano(res,
    lab = rownames(res),
    x = 'log2FoldChange',
    y = 'pvalue',
    xlim = c(-10, 10),
    ylim = c(0,30),
    title = 'zoom in',
    pCutoff = pCutoff,
    FCcutoff = 2,
    pointSize = 3.0,
    labSize = 3.0)
```

### Plot counts for lowest padj
```{r plot_counts, message=FALSE, warning=FALSE}
d <- plotCounts(dds, gene=which.min(res$padj), intgroup= c("condition", "group"), returnData=TRUE)
ggplot(d, aes(x=condition, y=count, shape=condition, col=group)) +
              geom_point(position=position_jitter(w=0.1,h=0.1))
```

### Plot counts for highest l2fc
```{r plot_counts_fc, message=FALSE, warning=FALSE}
d <- plotCounts(dds, gene=which.max(abs(res$log2FoldChange)), intgroup= c("condition", "group"), returnData=TRUE)
ggplot(d, aes(x=condition, y=count, shape=condition, col=group)) +
              geom_point(position=position_jitter(w=0.1,h=0.1))
```

### 100 most significant genes
```{r top_genes, message=FALSE, warning=FALSE}
resSignificant <- res[which(abs(res$log2FoldChange) > 1),]
resSignificant <- resSignificant[which(resSignificant$padj < 0.05),]
resSignificant <- resSignificant[order(resSignificant$padj),][1:100,]
resSignificant$gene_name <- ensembl.genes[rownames(resSignificant), ]$gene_name
resSignificant$chrom <- seqnames(ensembl.genes[rownames(resSignificant),])
resSignificant$start <- start(ensembl.genes[rownames(resSignificant),])
resSignificant$end <- end(ensembl.genes[rownames(resSignificant),])
```
```{r print_genes, message=FALSE, warning=FALSE}
library(kableExtra)
kable(resSignificant, "html") %>% kable_styling("striped") %>% scroll_box(width = "100%")
```

```{r write_genes, message=FALSE, warning=FALSE}
write.table(resSignificant, file=out.list, quote=FALSE, sep="\t", row.names=T)
```
